manuscript_notes

### 2024-10-30 notes/manuscript_notes.Rmd
update of notes.
notes/feedback on the manuscript

-> still to do.

@@

>>>>

-> I added a content list at the start of each Tutorial_x.Rmd file.

@@

# October 29, 2024 #

## minor comments ##

- I removed the repetition of the title on the intro page. This always annoys me!

- Minor point. Response time vs reaction time? Line 1 of the intro, we say response. 
But don't we want to say reaction time, as most people use that term?

- Makeham reference. Shouldn't it just be Makeham, 1860, rather than first and middle names?

- I would explicitly insert a reference in the main text for papaja(). See the citation subsection of this webpage: https://github.com/crsh/papaja. And see page 6 in the manuscript for the text that I added. I feel it is worth highlighting / advertising since the developers can get more recognition and maybe others will start using it too.

- I would remove reference to specific equations throughout the manuscript. Instead, I would just say it once, like you do at the start of section 2. e.g., we only supply regression equations in supps. then don't mention it again because it disrupts the flow in my view. Folks who want more can easily flick to the supps and read all they want.

- section 2.1. we use an abbreviation for EHA. and we also define it as hazard. Since we have already given the other terms for EHA, let's just use EHA. I hope that makes sense. And shouldn't we abbreviate to EHA the first time we mention it, rather than in section 2.1?

- section 2.1. You say...

>Continuous RT data is treated here as interval-censored data.

But in the context of this paragraph, this makes no sense and just seems tacked on and disconnected. Can we explain or remove?

- Also in section 2.1, you say in the middle of a paragraph...

>This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before.

Two things here. 1) At this stage in proceedings, I'm not even sure we need to give the benefits of using a hazard, but instead, just give the definition. So we might just remove it.
2) If we do include it, then this sentence does not communicate how or why it is more diagnostic, just that it is more diagnostic. And we provide no support for this claim. Later on in the discussion (I think) there is more detail regarding the greater diagnosticity. But here, we get caught between two posts and don't provide sufficient detail. So, I would just remove it from here for now, to make things shorter and simpler, at least for the time being.

- Also in section 2.1, you say ...

>As explained in part A of the supplementary material, the survivor function provides a context for interpreting the hazard function.

Again, I'm not sure we can leave people hanging like this. Instead, I would maybe wrap some of this contextualisation into the figure legend of Figure 2. And delete this sentence. Or include 1 or 2 sentences that explain how the survivor can help contextualise the hazard, but make it clear and relevant.

[[As I went back to the manuscript, I saw an old paragrpah and re-inlcuded it, as I think it is relevant. e.g., if we make the hazard and survivor functions a big deal by inlcuding them in a figure, then we really should explain what to make of them and why they are important.]]

- section 4.1.2. I don't understand why it is -187ms. This needs explaining because it is not obvious.

- I really like it when you mention the use of "our" custom functions on page 19. That seems exactly what we should be doing. i.e., present a workflow and functions that make life easier for others to do it.

- what does the following mean?

>duplicate terms in the model formula are ignored

This should be clearly explained or removed.

- I like the walkthrough of the parameter estimates in section 4.3.8. That's nice and clear.

- can we say why we have the following sentence?

>For illustration purposes, we only fitted the effects from model M3 (see Tutorial 2a) using the function glmer() from the R package lme4.

Were they too complex to fit or something else?

## regressions references ##

Gelman, A., Hill, J., & Vehtari, A. (2021). Regression and Other Stories. Cambridge University Press.

Winter, B. (2019). Statistics for Linguists: An Introduction Using R. Routledge.

## table 1 and table 2 ##

Personally, I think having Tables 1 and 2 separately is fine. Let's leave it for now.

## figure 1 ##

I changed the layout of figure 1 as I thought it was easier to see.
I also changed the way RT data were generated to a uniform distribution.

## figure 3 ##

I tried to insert a little more explanation as to why the joint consideration of hazard plus ca is psychologically interesting.
One problem I had, however, is that if I calculate mean average RT from these hazards, then there is no difference in RT. I guess if we want to visually illustrate the differences it might be nice to have an insert plot showing the typical interferece measure of mean RT and then the distribution plots.
However, it might also be better just to keep this super simple and make the point that we do at the moment, which is that both together help guide inferences. And let's worry less about plotting mean RT in this figure.

## life tables ##

section 4.1.2
We mention life tables, but have we defined them before this point? If not, we need to define them first. And mention why they are important. If not, they just come out of nowhere.

indeed, maybe the below quoted text should come earlier to explain life tables

>Table 3 shows the life table for condition "blank" (no prime stimulus presented) for participant 6. A life table includes for each time bin, the risk set (i.e., the number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), P(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.

## figure 5 ##

Move this to the Tutorial. It is way too complicated for an already long paper.
Folks who are interested can look at the tutorial.

## code can be more compact ##

e.g., this

```{r}
# load person-trial-bin oriented data set
ptb_data <- read_csv("../Tutorial_1_descriptive_stats/data/inputfile_hazard_modeling.csv")
# select analysis time window: (200,600] with 10 bins (time bin ranks 6 to 15)
ptb_data <- ptb_data %>% filter(period > 5)
# create factor condition, with "blank" as the reference level
ptb_data <- ptb_data %>% mutate(condition = factor(condition, labels = c("blank", "congruent","incongruent")))
# center TIME (variable period) on bin 9, and variable trial on number 1000 and rescale; Add dummy variables for each bin.
ptb_data <- ptb_data %>% 
        mutate(period_9 = period - 9,
               trial_c = (trial - 1000)/1000,
               d6  = if_else(period == 6, 1, 0),
               d7  = if_else(period == 7, 1, 0),
               d8  = if_else(period == 8, 1, 0),
               d9  = if_else(period == 9, 1, 0),
               d10 = if_else(period == 10, 1, 0),
               d11 = if_else(period == 11, 1, 0),
               d12 = if_else(period == 12, 1, 0),
               d13 = if_else(period == 13, 1, 0),
               d14 = if_else(period == 14, 1, 0),
               d15 = if_else(period == 15, 1, 0))
```

can instead be...

```{r}
# load person-trial-bin oriented data set
ptb_data <- read_csv("../Tutorial_1_descriptive_stats/data/inputfile_hazard_modeling.csv") %>% 
# select analysis time window: (200,600] with 10 bins (time bin ranks 6 to 15)
filter(period > 5) %>% 
# create factor condition, with "blank" as the reference level
mutate(condition = factor(condition, labels = c("blank", "congruent","incongruent"))) %>% 
# center TIME (variable period) on bin 9, and variable trial on number 1000 and rescale; Add dummy variables for each bin.
mutate(period_9 = period - 9,
       trial_c = (trial - 1000)/1000,
       d6  = if_else(period == 6, 1, 0),
       d7  = if_else(period == 7, 1, 0),
       d8  = if_else(period == 8, 1, 0),
       d9  = if_else(period == 9, 1, 0),
       d10 = if_else(period == 10, 1, 0),
       d11 = if_else(period == 11, 1, 0),
       d12 = if_else(period == 12, 1, 0),
       d13 = if_else(period == 13, 1, 0),
       d14 = if_else(period == 14, 1, 0),
       d15 = if_else(period == 15, 1, 0))
```

And I also wonder. Should this even be in the main text, given that it is in the tutorial?
I'm not sure. But it doesn't strike me as overly important to have the repetition or redundancy, and we have a long paper already. So folks might just get tired of having to figure out more stuff when they don't need to if they only want to read along. What do you think?

## model building ##

adapt_delta seems high. I rarely go above 0.99.
And why not init_r = 0.1 or similar instead init = aexactly zero every time.
0.1 means that the starting point varies a bit between -0.1 and 0.1 (i think)

## table 4 and 5 ##

These are way too long. 
My suggestions as follows:

1. just show con and inc 500 for the purposes of the manuscript. then in the main text and legend we can explain that it is just a selection of data for the purposes of understanding.
The full tables can be in SUpps if you feel it is necessary.

2. remove the .width column. just say in the legend 95% quantiles.

3. split c500 and i500 into two columns. e.g, trial =500, condition = con or inc but in two separate columns.

4. Round to two dp.

5. And what's going on with some of those hazard values and odds ratio values? Is it just because one value could be so small that the ratio can be absolutely massive? It just seems odd without any explanation? e.g., take a look at the odds ratio column in table 5.

For now, I just removed them to make the doc easier to manage.

## model formulas ##

some things are not clear, such as:

- model_M1. 
--The use of 0 and Intercept needs explaining.
--adapt_delta is high. I rarely have to go above 0.99. Does the model not look right if this is set lower? This could be a sign that the model is not a suitable structure.
--step_size - I have never set this. Can you send me some info on this parameter?
-- init set at exactly zero. This should be avoided whenever possible. Try init_r = 0.1. This means that the start points will vary between -0.1 and 0.1, rather than always being the same value (zero).
-- Although condition is defined earlier. I would make it clear in this section what condition means, as it is super important.

- model_M2
-- clearly define what period_9 means.
-- It seems to me that you have quadratic and 3rd order terms, but you only mention 3rd order in the text.

## loo plots ##

They seem to be missing from section 4.3.7. 
They would be useful in combinartion with model weights.

## psychological / cognitive interpretation ##

You say ...

>The fact that congruent primes generate negative effects for 80 ms (compared to no-prime trials) towards the end of the experiment, while incongruent primes generate negative effects for 120 ms throughout the experiment, suggests the involvement of separate cognitive processes.

Note - I removed the word "strongly" from here, as I'm just not sure it is appropriate and it adds nothing anyway.

And my question: We argue the above, but couldn't an equally plausible account be that it reflects the exact same cognitive process, but for different durations? i.e., different onsets and offsets? Is there some aspect to the argument that I am missing? If so, we need to provide those details or give the alternative interpretation. It is still interesting, but I think we need to be clearer here and a little more careful.

And a more general comment on this section, which I don't, as yet, have a perfect solution to.
On page 37, the paragraph starts with Panis and Schmidt (2016) distinguished between automatic ...
I like the aim of this paragraph because it brings folks back to cognitive mechanisms and cog psych, which many readers may be interested in, since that is who the article is aimed at.
In other words, it aims to show how all this fancy modelling and effort can bring folks to a meaningfully different interpretation of the operation of the underlying system.
With that said, after reading quite a bit of stats stuff, modelling stuff and code chunks, parameters etc., to them dump on top a bunch of new processes and terms that are not previously outlined, just feels too much for readers. Like they are punch-drunk at this point and we can tell them anything because they stopped listening or caring.

So, what to do? I outline some options below.

1. create a new subsection 4.3.9 Relevance for theory (or a similar title) where we place the paragraph and explain why and how such results matter for theory development. By doing so, we accept that it is long and complicated, but we think it is important enough to give it the space it needs.

2. We largely remove the paragraph and instead point readers towards Panis and Schmidt 2016. e.g., To see how such results matter for theory development and the distinction between competing theories, see Panis and Schmidt 2016.

## posterior predictions? ##

You say...

>To conclude this Tutorial 2a, Figure 7 shows the model-based hazard functions for each prime type for participant 6, in trials 500, 1000, and 1500.

Are these posterior predictions? If so, they need a little explanation.

## figure settings ##

The below figure settings are in the manuscript file.

```{r plot-settings}
## theme settings for ggplot
theme_set(
  theme_bw() +
    theme(text = element_text(size = 18), 
          title = element_text(size = 18),
          legend.position = "bottom")
)

## Set the amount of dodge in figures
pd <- position_dodge(0.7)
pd2 <- position_dodge(1)
```

This was confusing initially, because your figures definitely do not follow these settings.

Then I realised that you are just taking figures from files. 
So, the files used to generate the figures need this section adding to the top of the script, so that they are generated in a consistent manner.

Also, I would be explicit with sizing when saving figures, if you don't so this already.
If not, the size of the saved figure is determined by the plot window in RStudio and that can change depending on the screen and size of RStudio at the time.
And also set the figure dpi higher for a crisper resolution e.g., dpi=800..

https://ggplot2.tidyverse.org/reference/ggsave.html

e.g.,

```{r}
ggsave("sims/figures/haz_surv.jpeg",
       width = 7, height = 5, dpi=800)
```

One more important thing - colours/aesthetics etc. 
I think we can improve our data vis game. 
e.g., Figure 6 and 8 look great, but Figure 7 and 9 need some work.
I think we can pick better colour palettes e.g., https://r-graph-gallery.com/38-rcolorbrewers-palettes.html

I often go with the following palette ("Dark2"), see text and chunk below as an example. use colour and fill depending on whether you want the particular palette applied to colour or fill aesthetics or both.

  scale_colour_brewer(palette = "Dark2") + 
  scale_fill_brewer(palette = "Dark2") +

```{r}
p9.7 <-ggplot(data_hs, aes(x = timebin, y = probability,
                         colour = dv)) +
  geom_line(linewidth = 1) +
  geom_point(size=3) +
  scale_colour_brewer(palette = "Dark2") + 
  scale_x_continuous(breaks = 1:10) +
  scale_y_continuous(limits = c(0,1)) +
  ggtitle("hazard and survivor functions") 
p9.7
```

This is important for several reasons.
1. Data vis is just important and can really make a paper stand out.
2. We need to be consistent throughout and we don't want some tables/figures looking great and others looking half-finished or an after thought.

## figure legends ##

in some cases, much more information is needed. e.g., Figure 7 and 9. Folks need a guide.

## figure layout ##

have you considered 1 column instead of 3 for figure 7/9?
does it look better/easier to understand? Maybe try it and see?

Figure 9 looks odd. Can we re-szie and re-colour as sggested above. and maybe try one column for the facet rather than 3. To set this, just insert "ncol=1" into the facet argument. e.g., facet_wrap(~condition, ncol=1).

Figure 10 can be flipped such that the axes are opposite. e.g., flip x and y. then the text will be easier to read and plots easier to understand, in terms of numerical values.

## Discussion ##

I really liked reading the discussion. 
For the most part, it was compact and efficient and covered some big-ticket items.
The one exception was the 5.1 advantages section. It was all very clear, but it read like a background to the topic, which should come in the intro, rather than at the end. e.g., we've just been through all of this text, code, figures and model building and then we decide to say there are advantages?? It just seems out of sync.
So, I moved the text to Supplementary Material and created a new sub-section in supps.
There are several benefits of doing so.

1) we can keep the discussion focussed on things that are relevant to exp psych folks, rather than more general stats topics, which would be relevant for textbooks and the like.

2) we can link to these benefits earlier on when we give a background to EHA in section 2.1.
I've created a new sub-section in sectino 2 and added a sumamry paragraph that points the reader to the supplementary materials for the details.

3) We can add something else into the discussion that is more relevant to our aim here. I have tried to add in a section on common use-cases for exp psych folks. It might not quite work or fit in yet, but the motivation was something like this. You may want to use it if you are interested in the timeline of effects. But you may also need to use it from a counfound/technical/method perspective because it might qualify inferences made from mean-average comparisons. The reason is just to make it clear that you need not be interested in temporal states to find this useful on a methodological/analytical perspective.
